---
title: "Session 4 - HAVOK"
output: word_document
date: "2025-03-16"
---

```{r setup}
#| include: FALSE
knitr::opts_chunk$set(echo = TRUE)

for(packageName in c("deSolve",
                     "devtools",            # R packages needed here
                     "havok",
                     "here",
                     "magrittr",
                     "tidyverse")) {
  if(!is.element(packageName,               # If package is NOT installed...
                 installed.packages()[,1])) {
    install.packages(packageName)           #  ...then install it.
  }
  library(packageName,                      # Add package to environment
          character.only=TRUE,
          quietly=TRUE,
          verbose=FALSE)
}

here::i_am("Scripts/Session 4 - HAVOK.Rmd") # To help find all the files.

devtools::install_github("RobertGM111/havok") # Moulder et al. (2023)
library("havok")

options(show.signif.stars = FALSE)          # Don't stargaze
options(digits = 3)                         # Round to 3 digits by default
```

```{r data}
data("ECG_measurements")
data("Internet_Trolls")
```

# HAVOK

Data and code on the Moulder et al. site. Here is their Readme file:


<!-- **NOTE: This package is in active development (beta), updates regularly,
and may have bugs. Please report any bugs to the package maintainer.**

This package allows for modeling of chaotic systems as intermittently
forced linear systems through the use of Hankel Alternative View of
Koopman (HAVOK) analysis (Brunton, Brunton, Proctor, Kaiser, & Kutz,
2017). This package has additional functionality for the SINDy algorithm
(Brunton, Proctor, & Kutz, 2016), as well as other nonlinear analysis
methods. This package also contains functionality for model optimization
through parallel processing via the `phavok()` function. -->

## Lorenz attractor - Basic `havok()` Function Usage

Simulate data from a Lorenz attractor.

```{r}
#Generate Data
##Set Lorenz Parameters
parameters <- c(s = 10, r = 28, b = 8/3)
n <- 3
state <- c(X=-8, Y=8, Z=27) ##Inital Values

dt<-0.001
tspan<-seq(dt,200,dt)
N<-length(tspan)

Lorenz <- function(t, state, parameters) {
  with(as.list(c(state, parameters)), {
    dX <- s * (Y - X)
    dY <- X * (r - Z) - Y
    dZ <- X * Y - b * Z
    list(c(dX, dY, dZ))
  })
}

out <- ode(y = state, times = tspan, func = Lorenz, parms = parameters, rtol = 1e-12, atol = 1e-12)
xdat <- out[,"X"]
t <- out[,"time"]

# Run HAVOK Analysis
hav <- havok(xdat = xdat, dt = dt)
#> Warning in havok(xdat = xdat, dt = dt): Agument "devMethod" not selected.
#> Defaulting to devMethod = "FOCD"
```

To plot the resulting time series and forcing term use `plot(hav)`

```{r}
plot(hav, what = "both")
```

## Internet Trolls - Parameter Optimization via Parallel Processing

Load internet trolls data (Full data available at
[www.fivethirtyeight.com](https://fivethirtyeight.com/features/why-were-sharing-3-million-russian-troll-tweets/?fbclid=IwAR0PEu76Gtl_jHaXvghjmdQOAio_qgqe-bJp6ZQRbh5mt94EPQ884IA0XCA).
The original data file for this example contains data from June 19,
2015, to Dec. 31, 2017.

```{r}
data("Internet_Trolls")
```

Conduct parameter optimization via `phavok()`

troll activity extracted 4 times per day during the US presidential election year 2016 on 11 different topics 

```{r}

results.all.truncated[results.all.truncated$Type=="Right",] -> 
  right                                     # only right-wing trolls
right$Topic3 ->                             #  Russian Twitter troll posting 
  xdat                                      #  activity on the topic of Racial
                                            #  Justice/Black Lives Matter
0.25 ->                                     # 4 measurements per day
  dt

phavok(xdat = xdat,                         # All possible rs within specified
       dt = dt,                             #  stackmax range, no 
       stackmaxes = 28:58) ->               #  sparsification dimension
  results
```

```{r}
phav <- phavok(xdat = xdat, dt = dt, stackmaxes = 28:58, rs = 2:10, random = 0.5, sparsify = T, sparserandom = 0.4)
```


## References

- Brunton, S. L., Brunton, B. W., Proctor, J. L., Kaiser, E., &
  Kutz, J. N. (2017). Chaos as an intermittently forced linear system.
  Nature communications, 8(1), 19.

- Brunton, S. L., Proctor, J. L., & Kutz, J. N. (2016). Discovering
  governing equations from data by sparse identification of nonlinear
  dynamical systems. Proceedings of the National Academy of Sciences,
  113(15), 3932-3937.

- Moulder, R. G., Martynova, E., & Boker, S. M. (2021). Extracting
  nonlinear dynamics from psychological and behavioral time series
  through HAVOK analysis. Multivariate Behavioral Research, 1-25.
