---
title: "Session 8 - Infectious Diseases"
output: word_document
date: "2025-03-16"
---

```{r setup}
#| include: FALSE
knitr::opts_chunk$set(echo = TRUE)

for(packageName in c("devtools",            # R packages needed here
                     "here",
                     "magrittr",
                     "NlinTS",
                     "nonlinearAnalysis",
                     "rEDM",
                     "Rssa",
                     "tidyverse")) {
  if(!is.element(packageName,               # If package is NOT installed...
                 installed.packages()[,1])) {
    install.packages(packageName)           #  ...then install it.
  }
  library(packageName,                      # Add package to environment
          character.only=TRUE,
          quietly=TRUE,
          verbose=FALSE)
}

i_am("Scripts/Session 8 - Infectious Diseases.Rmd") # To help find all the files.

options(show.signif.stars = FALSE)          # Don't stargaze
options(digits = 3)                         # Round to 3 digits by default
```

```{r data}

```

First, we reconstructed a (shadow) delay state space.

We then used that delay state space to run a series of tests, each of which has some power to distinguish between one of the mimics and nonlinearity.

Tests for noise included the BDS test, various surrogate tests (using discriminant statistics such as MLCE, etc.) and various entropy tests.

Tests of linearity included various AR tests and nonlinear signal detection (via the Nash-Sutcliffe efficiency)

Tests of stationarity included nonlinear cross-prediction, where we segment the data and look for degradation in cross-prediction.

In many of these we used surrogate data to create confidence intervals.

If there was support for a nonlinear signal, we then used singular spectrum analysis to remove some noise and low-frequency components (a.k.a., trend), leaving us with a cleaner signal.

After cleaning the data, we repeated our tests. 

Finally, we used convergent cross mapping and transfer entropy to explore the extent to which different data streams were ”causally” (in quotes) related or if they were separable.


```{r testsRound1}

```

