---
title: "Session 5 - Python Setup"
output: word_document
date: "2025-03-16"
---


```{r setup}
#| include: FALSE
knitr::opts_chunk$set(echo = TRUE)

for(packageName in c("devtools",            # R packages needed here
                     "reticulate",
                     "tidyverse")) {
  if(!is.element(packageName,               # If package is NOT installed...
                 installed.packages()[,1])) {
    install.packages(packageName)           #  ...then install it.
  }
  library(packageName,                      # Add package to environment
          character.only=TRUE,
          quietly=TRUE,
          verbose=FALSE)
}
```

# Install Tensorflow

Requires Python. Let's follow https://tensorflow.rstudio.com/install/

First, do you have Python?
```{r}
system("python3 --version")
```
If you have Python, and the output is "Python 3.10.x" where "x" is some number, then you should be good to go.

If you need Python 3.10, remove the "#| eval: FALSE" line in the next chunk and run the chunk. This will take a few minutes. (package:reticulate connects RStudio and Python. See https://rstudio.github.io/reticulate/)
```{r}
#| eval: FALSE
reticulate::install_python(version = "3.10")
```


Install tensorflow stuff. If asked, update All.
```{r}
devtools::install_github("rstudio/tensorflow",
                         force = TRUE)
```
The install_github() command takes me about 10-15 seconds or so plus time for updating other packages if needed.


There are a few hundred lines of output from the next command. Don't worry, you don't need to know what any of it means.
```{r}
library(tensorflow)
tensorflow::install_tensorflow(envname = "r-tensorflow",
                               version = "2.16")
```

Now, confirm all is working fine:
```{r}
library(tensorflow)                         # Confirm tensorflow installation
tf$constant("Hello TensorFlow!")
```

If all is working as expected, the last command in the previous chunk will produce this output:

tf.Tensor(b'Hello TensorFlow!', shape=(), dtype=string)


If you got this far, congratulations! If not, then we'll work on it.

## Installing Python Packages

There are many Python packages that we may require. Hence, you should know how to install them. Here are a few common installs:
```{r}
reticulate::py_install(c("matplotlib",
                         "numpy",
                         "pandas",
                         "scikit-learn"))
```


# Some Introductory Machine Learning Sites

Here are some introductory machine learning/tensorflow sites, in case you want to try on your own:

https://tensorflow.rstudio.com/tutorials/quickstart/beginner
https://www.datacamp.com/tutorial/neural-network-models-r
https://www.r-bloggers.com/2021/04/deep-neural-network-in-r/#google_vignette



