---
title: "Session 8 - Infectious Diseases"
output: word_document
date: "2025-03-16"
---

```{r setup}
#| include: FALSE
knitr::opts_chunk$set(echo = TRUE)

for(packageName in c("here")) {
  if(!is.element(packageName,               # If package is NOT installed...
                 installed.packages()[,1])) {
    install.packages(packageName)           #  ...then install it.
  }
  library(packageName,                      # Add package to environment
          character.only=TRUE,
          quietly=TRUE,
          verbose=FALSE)
}

i_am("Scripts/Session 8 - Infectious Diseases.Rmd") # To help find all the files.

source(here("Scripts/Setup.R"))
```

```{r data}
load(here("Data/NY Scarlet Fever.RData"))

# Get stuff
```

```{r testsRound1}
bds.test()

```

Pick a stream, any stream.
```{r}
```

Choose a delay
```{r}
mi_ksg(MotorVehicle) ->
  mi

plot(mi)

which.troughs(mi)[1] - 1 ->                 # This automates the visual
  delay                                     #  we used. 0-indexed, so - 1

```

Embedding dimension
```{r}

```

Theiler window
```{r}

```
Embed
```{r}

```

Plot (if possible)
```{r}

```

## Tests

Noise: Is the signal simply noise?
```{r}
bds.test(x = ,                    # data stream
         m = ,                             # Embed. dim. of 2, 3, and 4
         eps = seq(0.1 * sd(MotorVehicle),    # How close is close? Try various
                   2 * sd(MotorVehicle),      #  values.
                   length.out = 6)) ->
  bds                                       # H0: Series of random i.i.d.
bds$p.value                                 # For various dimensions and
                                            #  definitions of "close"
```
If $p < 0.05$ there's a signal. (The time graph indicates that the signal probably isn't linear.)

Maximum characteristic Lyapunov exponent
```{r}
maxLyapunov(MotorVehicle,
            max.embedding.dim = 6,
            radius = 5)
```
Why do we care about chaos? Chaos in real data becomes almost impossible to model.


Nonlinearity: Is there a nonlinear signal?
```{r}
nse(MotorVehicle)$nse               # Nash-Sutcleffe coefficient of
                                            #  efficiency

```
Typically, nse > 0.65 is proposed as the threshold for a nonlinear model to exist. nse > 0 means the nonlinear model is a better predictor than the mean of the data; nse < 0 inidcates that the mean is a better predictor than the model. Of course, these are dependent on embedding dimensions, delays, and so on, so this will not always produce what is desired.

Test ARIMA models
```{r}
Keenan.test(MotorVehicle)
tlrt(MotorVehicle, p = 26)
Tsay.test(MotorVehicle, order = 26)            # H0: Linear AR; H1: Quadratic AR
```

Nonstationary test: Is the signal nonstationary?
```{r}
nl_cross_pred(MotorVehicle,
              nseg = 5,                     # Default is 5
              m.max = 6)                    # Default is 6

```

```{r}
crqa()

```


There are other tests as well

## Is There a Signal?

Surrogate data testing
```{r}
surrogateTest(lh_df$Hare[1:39])
```


CCM
```{r}
rEDM::ccm()
```

# Model

## SINDy

## DMD

## Extreme Value Statistics