---
title: "Session 1 - Derivatives"
output: word_document
date: "2025-03-16"
---

STILL TO DO:
- regularized derivatives in R (not Python)
- plot regularized derivatives

```{r setup}
#| include: FALSE
knitr::opts_chunk$set(echo = TRUE)

for(packageName in c("devtools",            # R packages needed here
                     "doremi",
                     "here",
                     "magrittr",
                     "NlinTS",
                     "nonlinearAnalysis",
                     "rEDM",
                     "Rssa",
                     "tidyverse")) {
  if(!is.element(packageName,               # If package is NOT installed...
                 installed.packages()[,1])) {
    install.packages(packageName)           #  ...then install it.
  }
  library(packageName,                      # Add package to environment
          character.only=TRUE,
          quietly=TRUE,
          verbose=FALSE)
}

i_am("Scripts/Session 1 - Derivatives.Rmd") # To help find all the files.

install_github("barneyricca/ndstools")
library(ndstools)

options(show.signif.stars = FALSE)          # Don't stargaze
options(digits = 3)                         # Round to 3 digits by default
```

```{r data}
read.csv(here("Data/Mayport_raw.csv"))[,1] ->
  mp
load(here("Data/HRF.RData"))
```

First, we can look at the Mayport data
# Finite differences

## Basic

```{r finiteDifferences}
data.frame(mpv = diff(mp),
           time = seq(from = 1.5,
                      to = length(mp) - 0.5,
                      by = 1)) ->
  mpv_fd
```

## Local Linear Analysis (LLA)
Set the time step to 1/2 and you get basic finite differences. For integer time steps, you can do LLA; you will lose a number of data points equal to $\tau$ at each end of the sequence, so there's a trade-off
```{r}
1 -> tau

# A padding trick; it makes things easier
c(rep(NA, tau), mp[1:(length(mp) - tau)]) -> mpmt
c(mp[(tau+1):length(mp)], rep(NA, tau)) -> mppt

(mppt - mpmt) / (2 * tau) ->                # Because of the padding, this is
  mpv_lla                                   # Pretty much the LLA approach

```

## Higher order

HBR like 4th order. This is sort of a generalization of the LLA approach. This keeps the times the same, although we lose two points at either end.

```{r 4thOrderFiniteDifferences}

# Pad each end; it makes things easier below
c(NA, NA, mp[1:(length(mp)-2)]) -> mpm2
c(NA, mp[1:(length(mp)-1)]) -> mpm1
c(mp[2:length(mp)], NA) -> mpp1
c(mp[3:length(mp)], NA, NA) -> mpp2

# This is 1 unless there are times as well as levels in the data. (I.e., if
#  there are only indices, leave this as 1 -> delta. Otherwise, use the
#  sampling period.)
1 -> delta

# 4th order centered finite difference
(8*(mpp1 - mpm1) - mpp2 + mpm2) / 12 ->
  mpv_4fd

```

You can also use an approach similar to *Savitzky-Golay* filtering (known to statisticians as *loess smoothing*) to generalize the higher order approach. This also shows up as the basis for the functional data analysis approach.

## Generalized Local Linear Analysis (GLLA)

```{r glla}
doremi::calculate.glla(mp,
                       time = 1:length(mp)) ->
  mpv_glla
```

# Functional data analysis (FDA)

Akin to loess smoothing
```{r glla}
doremi::calculate.fda(mp,
                      time = 1:length(mp),
                      spar = NULL) ->
  mpv_fda
```

# Regularized Derivatives

# Empirical Bayes

Can use irregular time intervals (i.e., work with missing data!).
```{r emb}
#| eval: FALSE
EMBestimates(DataM = mp,
             Nobs = mp_dss$dim,
             maxorder = 2,
             dt = 1) ->                     # Or provide TimeM for irregular
  mpv_emb                                   #  data spacing
```

# Multilevel-based model: Generalized Orthogonal Local Derivatives (GOLD)

```{r glla}
doremi::calculate.gold(mp,
                       time = 1:length(mp)) ->
  mpv_gold
```

# (Visual) Comparisons

```{r}
par()$mfrow ->                              # Want to restore all the graphics
  row.keep                                  #  parameters afterward
par()$mar ->
  mar.keep

{                                           # Finite differences
  par(mfrow = c(4,1))
  par(mar = c(0,4,2,0))                     # Avoid "margins too large" error
  plot(mp, type = 'l', ylab = "Mayport")
  plot(mpv_fd$time, mpv_fd$mpv, type = 'l', ylab = "FD")
  plot(mpv_lla, type = 'l', ylab = "LLA")
  plot(mpv_4fd, type = 'l', ylab = "4th order")
}

{
  par(mfrow = c(3,1))
  par(mar = c(0,4,2,0))
  plot(mp, ylab = "Mayport",
       type = 'l')
  plot(mpv_fda$dsignal[,1], ylab = "FDA Level",
       type = 'l')
  plot(mpv_fda$dsignal[,2], ylab = "FDA Velocity",
       type = 'l')
}

{
  par(mfrow = c(3,1))
  par(mar = c(0,4,2,0))
  plot(mp, ylab = "Mayport",
       type = 'l')
  plot(mpv_glla$dsignal[,1], ylab = "GLLA Level",
       type = 'l')
  plot(mpv_glla$dsignal[,2], ylab = "GLLA Velocity",
       type = 'l')
}

# Still to do:
# Regularized

{
  par(mfrow = c(3,1))
  par(mar = c(0,4,2,0))
  plot(mp, ylab = "Mayport",
       type = 'l')
  plot(mpv_gold$dsignal[,1], ylab = "GOLD Level",
       type = 'l')
  plot(mpv_gold$dsignal[,2], ylab = "GOLD Velocity",
       type = 'l')
}

{
  par(mfrow = c(3,1))
  par(mar = c(0,4,2,0))
  plot(mp, ylab = "Mayport",
       type = 'l')
  plot(mpv_emb$Derivatives[,1], ylab = "EMB Level",
       type = 'l')
  plot(mpv_emb$Derivatives[,2], ylab = "EMB Velocity",
       type = 'l')
}

par(mfrow = row.keep)                       # Restore graphics layout
par(mar = mar.keep)

```

