---
title: "Session 1 - MLM"
output: word_document
date: "2025-03-16"
---

```{r setup}
#| include: FALSE
knitr::opts_chunk$set(echo = TRUE)

for(packageName in c("here",
                     "lmerTest",            # lme4 package with tests
                     "magrittr",
                     "mlmhelpr",
                     "nlme",
                     "tidyverse")) {
  if(!is.element(packageName,               # If package is NOT installed...
                 installed.packages()[,1])) {
    install.packages(packageName)           #  ...then install it.
  }
  library(packageName,                      # Add package to environment
          character.only=TRUE,
          quietly=TRUE,
          verbose=FALSE)
}

i_am("Scripts/Session 1 - MLM.Rmd") # To help find all the files.

options(show.signif.stars = FALSE)          # Don't stargaze
options(digits = 3)                         # Round to 3 digits by default
```

# Simulated Data
```{r}
load(here("Data/Simpson.RData"))
```


# Radon Modeling

```{r data}
read.delim(here("Data/cty.dat"),            # From Gelman & Hill; radon data.
           sep = ',') -> U_df               #  Has Uranium counts
read.delim(here("Data/srrs2.dat"),          # 
           sep = ',') -> srrs_df

```

```{r wrangle}
"ctfips" -> 
  colnames(srrs_df)[24]
left_join(srrs_df,                          # Connect Uranium data to activity
          U_df,
          by = c("stfips","ctfips")) %>%    # State and county match
  filter(state == "MN") %>%                 # Use only Minnesota
  select(idnum, floor, activity, Uppm,      # Floor is important
         ctfips, county) %>%
  mutate(rad = ifelse(activity == 0,        # G & H suggest logarithms, but
                      log(0.1),             #  there are 4 @ zeroes.
                      log(activity))) ->    # G & H suggest logarithm
  radon_df

```

```{r notPooled}
radon_df %$%
  lm(activity ~ Uppm + floor + county) ->
  rad_no_pool

summary(rad_no_pool)
```

```{r completePooling}
radon_df %$%
  lm(activity ~ Uppm + floor) ->
  rad_complete_pool

summary(rad_complete_pool)
```

```{r partialPooling}
radon_df %$%
  lmer(activity ~ Uppm + floor +
         (1 + floor | county)) ->
  rad_mlm

summary(rad_mlm)
```

The intraclass correlation (ICC) tells us more-or-less how important the nesting is. Really, ICC is only good for "random-intercept" models, not "random-slope" models. Still, values "near" 0 indicate nesting not very important.
```{r ICC}
?icc                                        # Note the warning!
mlmhelpr::icc(rad_mlm)                      #

```


# Rats' Diets

Body weight of many rats on different diets, measured over several weeks. There are 3 levels; see https://rpsychologist.com/r-guide-longitudinal-lme-lmer#partially-nested-models or https://www.alexanderdemos.org/Mixed5.html for more information about doing this.

```{r}
data("BodyWeight")                          # From package:nlme



BodyWeight %$%
  lmer(weight ~ Time +                      # weight depends upon Time nested
         (Time | Rat / Diet)) ->            #  in Rat nested within Diet
  w1                                        # Ignore convergence issues for
                                            #  purposes of this demo.
summary(w1)                                 # Time, Time | Rat, and Time | Rat
                                            #  | Diet all "significant."
```

