---
title: "Session 2 - DSA"
output: word_document
date: "2025-03-16"
---

```{r setup}
#| include: FALSE
knitr::opts_chunk$set(echo = TRUE)

for(packageName in c("devtools",            # R packages needed here
                     "doremi",
                     "haven",
                     "here",
                     "lmertest",
                     "magrittr",
                     "tidyverse")) {
  if(!is.element(packageName,               # If package is NOT installed...
                 installed.packages()[,1])) {
    install.packages(packageName)           #  ...then install it.
  }
  library(packageName,                      # Add package to environment
          character.only=TRUE,
          quietly=TRUE,
          verbose=FALSE)
}

i_am("Scripts/Session 2 - DSA.Rmd")         # To help find all the files.

options(show.signif.stars = FALSE)          # Don't stargaze
options(digits = 3)                         # Round to 3 digits by default
```

# Example

Estimate from some linear oscillator data? Maybe?

# Montpetit et al., (2010)

Data and process from: Montpetit, M. A., Bergeman, C. S., Deboeck, P. R., Tiberio, S. S., & Boker, S. M. (2010). Resilience-as-process: Negative affect, stress, and coupled dynamical systems. Psychology and Aging, 25(3), 631–640. https://doi.org/10.1037/a0019268

```{r data}
#| include: FALSE
read_sas(here("Data/level1.sas7bdat")) ->   # Data from Montpetit et al.
  level1_df                                 #  EMA, already totaled.
read_sas(here("Data/level2.sas7bdat")) ->   # Survey data
  level2_df

```

```{r derivatives}
# Montpetit used GLLA. Do this idiographically for NAffect and stress.

level1_df %>%
  group_by(id) %>%
  mutate(NA0 = calculate.gold(NAffect,
                              time = day)$dsignal[1]) %>%
  mutate(NA1 = calculate.gold(NAffect,
                              time = day)$dsignal[2]) %>%
  mutate(NA2 = calculate.gold(NAffect,
                              time = day)$dsignal[3]) %>%
  mutate(S0 = calculate.gold(stress,
                              time = day)$dsignal[1]) %>%
  mutate(S1 = calculate.gold(stress,
                              time = day)$dsignal[2]) %>%
  mutate(S2 = calculate.gold(stress,
                              time = day)$dsignal[3]) %>%
  ungroup() ->
  gold_df

inner_join(gold_df,                         # Keep only id that appear in both
           level2_df,                       #  level1_df and level2_df. This is
           by = "id") ->                    #  different than Montpetit.
  mont_df

str(mont_df)
```

Key:
- id
- PAffect: sum of 10 PANAS positive affect items
- NAffect: sum of 10 PANAS negative affect items
- stress: sum of Perceived Stress Scale (items 1, 2, 3, 6, 7, 8, 9, 10, 11, 14)
- dispres (dispositional resilience) 
- frndsup (friend support)
- famsup (family support)

```{r mlm}
mont_df %$%
  lmer(S2 ~ S0 + S1 + NA2 +
         (1 + dispres + frndsup + famsup | id)) ->
  dum1
```

